flowCore: A Bioconductor software package for high throughput flow cytometry analysis

### 2008-08-15 paper/manuscript.tex
I don't believe I'll be prepared to cite plateCore by the time the paper is submitted.
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@33211 bc3139a8-67e5-0310-9ffc-ced21a209358
higher resolution for the flowQ image
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@33225 bc3139a8-67e5-0310-9ffc-ced21a209358
changes suggested by Ryan and Josef
git-svn-id: https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@33253 bc3139a8-67e5-0310-9ffc-ced21a209358

\subsubsection*{Results}
The \Rpackage{flowCore} package provides the FCM community a shared
research platform that enables bioinformaticians, computer scientists,
and statisticians to work collaboratively with biologists and
clinicians to develop novel methods for FCM data analysis.
\subsubsection*{Conclusions}
We hope that our framework will be the foundation for fruitful 
collaborative research involving multiple scientific fields and
that it will help facilitate the
development and deployment of FC-HCS methods to complex 
scientific and clinical applications.

@@

Some of the consequences of this lag in development are difficulties
in maintaining the integrity and documentation of extremely large
datasets, assessing measurement quality, developing validated assays,
controlling the accuracy of gating techniques, automating complex
gating strategies, and aggregating statistical results across large
study sets for further analysis.  Among the most serious consequences
of the current situation is, however, that it is very difficult for
new analysis approaches to find there way into standard practice. We
believe that this barrier to the development and dissemination of new
analysis methods is one of the fundamental restraints on the future
expansion of FC-HCS in both clinical and research applications.

@@

The computational tools we have developed are distributed in the R
software language as the Bioconductor package
\Rpackage{flowCore}. \Rpackage{flowCore} is a freely available, highly
functional and extensible FCM data analysis platform that enables
researchers to efficiently handle FC-HCS data and encourages open
development of tools for their coherent analysis. Among the issues
addressed by \Rpackage{flowCore} are computationally efficient data
structures and a range of specialized methods for compensation,
transformation, and gating.  \Rpackage{flowCore} runs on Windows, Mac
OS X and Linux/Unix operating systems. Conceptually, the underlying
data structures that build the core of our software can be implemented
in any programming language, and we chose the R language for its
advantages in statistical programming and visualization. In this
paper, we describe the functionality and methods underlying
\Rpackage{flowCore}, and provide examples of their use.

@@

\subsection*{Existing data standards and conventions}
Traditionally, the majority of FCM experiments have been analyzed by
manual data inspection in one or two dimensions, or by very basic
comparisons of summary statistics and most of the currently available
analysis tools are designed to reflect this workflow.  We believe that
these approaches, in addition to being expensive and labor intensive,
do not fully address the highly complex nature of FCM data; in
particular, they disregard many of the fundamental aspects of the
data, such as its underlying distribution and its high-dimensional
nature. Furthermore, the subjective character of manual analyses are a
major obstacle to reproducibility. For FC-HCS data, unassisted manual
inspection is extremely time consuming, and robust statistical methods
need to be developed to point investigators to interesting aspects of
the data, or to potential problems. While the expert knowledge of
immunologists and researchers remains crucial for the understanding of
FCM data, we believe that collaboration with other research fields
such as statistics and computer science can greatly improve the
relevance of FCM in today's high-throughput paradigm.

@@

The notion of classes from an object-oriented programming language
provides one coherent way to describe these richer data structures,
and functions or methods that work on them allow for interaction and
manipulation. Many of the currently available software solutions offer
only limited support for such self-contained structures, or make use
of containers like XML or binary storage formats that are designed
specifically for particular user interfaces and hence are not easily
amenable to programmatical access. In addition, the closed-source
nature of these products often makes it impractical to integrate them
into analysis pipelines. In this manuscript we describe classes for
FCM data analysis and their implementation in R, however, they could
just as easily be implemented in any other language (e.g., Java,
C++). Software written in those languages could use similar data
structures, thereby simplifying communication and the interchange of
data between analysis tools.

The \Rpackage{flowCore} framework presented here can import and
process raw data FCS files along with their complete set of
file-specific metadata (Figure~\ref{fig1:FrameWork}).  Moreover, it is
a software implementation of the Gating-ML CR, an new flow
cytometry data standard developed in collaboration with the ISAC Data
Standards Task Force, which makes it possibile to integrate
\Rpackage{flowCore} in existing work flows and to communicate with any
other FCM tool that adheres to the proposed standard\citep{SpidlenInPressCytometryA}. Adherence to
standards also plays a critical role in the ability of new methods
based on \Rpackage{flowCore} to find their way back into the standard
practices for FCM data analysis. \Rpackage{flowCore} is not a
GUI-driven software, and all operations are done using a command line
interface.  It is possible to add a more elaborate user interface on
top of this infrastructure, however the focus in this paper is on a
programmatic approach to enable the convenient development of novel
analysis methods and automation of complex analysis approaches.  By
taking the burden of data management from the programmer, and by
providing well-defined programming interfaces (APIs), it is possible
to readily test new ideas and to easily extend the framework's
functionality.

@@

\Rpackage{flowCore}'s primary task is the representation and basic
manipulation of FCM data. This is accomplished through a data model
very close to that adopted by other Bioconductor packages. All
information from a single FCS file, i.e., the collection of events and
the accompanying metadata, is stored in one single container. We call
the structure that hold this data a \Rclass{flowFrame}. Raw data
values as well as associated metadata of a \Rclass{flowFrame} can be
accessed programmatically. Most commonly, the metadata consist of
descriptors of the stains used in the experiment and the respective
measurement channels, information about compensation performed at the
instrument side and any additional keywords the user deems to be
important to annotate the data. A number of quality checks are
performed during the creation of a \Rclass{flowFrame} to ensure data
integrity.

\subsubsection*{flowSet}
In high-throughput FCM, many of the analysis tasks need to be
performed consistently across multiple flow measurements, hence we
introduce the concept of a collection of \Rclass{flowFrames} called a
\Rclass{flowSet}. A \Rclass{flowSet} is a container for multiple
\Rclass{flowFrames} along with relevant information associated with
each individual frame such as descriptions of the cell sample, the
treatment to which the sample was subjected, or the location of that
sample in a 96 or 384 well plate.  \Rclass{flowSets} manage the
consistent application of operations on the individual
\Rclass{flowFrames} and shift the burden of keeping score of the
metadata from the user to the infrastructure, thus reducing the risk
of errors (\textit{e.g.}, mixups of sample labels). The
\Rclass{flowSet} structure can be readily extended to incorporate the
potentially complex metadata associated with even larger FC-HCS
experiments such as clinical trials where hundreds patients might
provide samples at different points in time over the course of the
experiment.

%%Figure1 Diagram of the data structures and the basic operations
\begin{figure}
\centering
\includegraphics[width=0.9\textwidth]{Figure1-flowCoreFrameWork.jpg}
\caption{\label{fig1:FrameWork}{For each experiment, FCS files (as
    individual \Rclass{flowFrames}), phenotypic and metadata are
    stored in a \Rclass{flowSet}. Each \Rclass{flowFrame} in a
    \Rclass{flowSet} corresponds to one FCS file. All basic operations
    (e.g., compensation, transformation, gating) can be applied to
    either single \Rclass{flowFrames} or a \Rclass{flowSet}
    simultaneously.}}
\end{figure}

@@

The basic operations in FCM analyses are typically the same: the data
need to be compensated (if that was not already done on the
instrument) and transformed, and sup-populations of interest need to
be selected based on a set of (predominantly sequential) gates. All
software solutions for FCM analysis offer support for these
operations, typically in an interactive, GUI-driven interface. The
approach we have taken in \Rpackage{flowCore} is to abstractly
describe these operations and build a set of tools to perform them on
both \Rclass{flowFrames} and \Rclass{flowSets}. While transformation,
and to a certain extent compensation, are fairly routine operations
with only limited potential for improvement, being able to implement
new methodologies for gating of FCM data, and extending the
capabilities of \Rpackage{flowCore} through object oriented
programming are features that clearly sets our framework apart from
other FCM analysis tools. By factoring out as much of the bookkeeping
as possible, programmers can focus on the actual operations rather
then having to deal with the tedious details of data integration and
access. Third-party methods can act on their own as first-class
citizens in the analysis framework, without breaking the work flow or
the basic infrastructure. This design allows for the straightforward
extension of \Rpackage{flowCore}'s capabilities, and has already
fostered the development of a number of valuable add-ons
\citep{lo2008agf,sarkar2008ufv} \textit{(Fixme Errol: Will plateCore
  be in Bioconductor soon. It should be if we want to cite it.)}.

@@

Data transformation is essential for both data visualization and
modelling \citep{lo2008agf}. The major transformations that are
routinely used in FCM analysis have been implemented in
\Rpackage{flowCore} (\textit{e.g.,} quadratic, log or arcsinh, see
Table~\ref{table1} for a complete list). Furthermore, the design of
the R language makes it easy to define arbitrary functions to apply to
the data of individual \Rclass{flowFrames} or entire
\Rclass{flowSets}, respectively. Compensation is available for both
\Rclass{flowFrames} and \Rclass{flowSets}, and the software also
offers functionality to compute spillover or compensation matrices
from a set of appropriate compensation samples.

\begin{table}[ht]
\begin{center}
\begin{tabular}{|l|l|}
\hline
\multicolumn{2}{|c|}{Data Transformations} \\
\hline
linear & $ax + b$ \\
quadratic & $ax^2 + bx + c$ \\
natural logarithm & $ln(x)(r/d)$ \\
logarithm & $log(x)(r/d)$ \\
biexponential & $root(a*e^{(b*y)}-c*e^{(-d+y)}+f)$ \\
logicle& $root(Te^{-(m-w)}*(e^{(y-w)}-p^2e^{-(y-w)/p}+p^2-1)$ \\
truncate & $x_{x<=a} = a$ \\
scale & $(x-a)/(b-a)$ \\
split-scale & combination of linear and logarithm \\
arcsinh & $asinh(a + bx)+c$ \\
\hline
\end{tabular}
\caption{\label{table1}Data transformations implemented in
  \Rpackage{flowCore}.}
\end{center}
\end{table}

\subsubsection*{Gating}
%%FIXME: I would make more of the filters and possibly try to make it
%%clearer how much stuff thre is for composition of gates/filters etc
%%
%% somewhere notions of apply and summary methods should appear and if
%% we have anything figured out yet on how to do the subsetting etc
%% that too would go nicely in here (here I am thinking of our
%% discussion of the sort of workflow stuff)

In \Rpackage{flowCore}, gating operations are represented by classes
that can be extended in an object-oriented manner
(Table~\ref{table2}). Basic gate types such as rectangular gates,
ellipses and polygon gates are implemented as part of the
framework. In addition, we introduce the notion of data-driven gates,
or filters, for which the necessary parameters are computed based on
the properties of the underlying data, for instance by modelling data
distribution or by density estimation. This approach is fundamentally
different from the traditional application of static gating regions
across samples, as it is able to take into accounts unforeseen changes
in signal intensities (\textit{e.g.,} drifts in the instrumentation
over time or sample variablity).

@@

Gating, as well as all other operations in flowCore, can be applied
over each individual frame in a \Rclass{flowSet}, and summary methods
provide information about the outcome of these operations. In
addition, the result of a gating operation can be used to subset the
input \Rclass{flowFrame} or \Rclass{flowSet}, either by filtering out
negative events or by splitting in multiple subpopulations. This
design allows to easily combine all of \Rpackage{flowCore}'s
components into complex work flows.

@@

In addition to the \Rpackage{flowCore} package that offers basic
infrastructure, we have implemented a range of additional Bioconductor
packages that are dedicated to more specific tasks of FCM data
analysis. The \Rpackage{flowViz} package \citep{sarkar2008ufv}
provides sophisticated data visualization tools, that make use off
multivariate trellis plotting \citep{lattice}.  These functions can be
used to quickly generate customized plots for extended cytometry data
sets for both direct data inspection and quality control.  The objects
metadata information can be used to arrange the layout and composition
of the plots.  Furthermore, the design and the API of the
visualization software is very generic, and users can readily extend
its capabilities by providing self-defined plotting functions.  The
\Rpackage{flowQ} package offers more advanced quality assurance (QA)
methodology and a framework to create interactive web-based reports of
QA results.  Most of \Rpackage{flowUtil}'s content deals with data
import and export, implementing the flow-cytometry specific standard
markup language.  Finally, \Rpackage{flowStats} provides more
elaborate statistical methods that are relevant in the context of flow
cytometry data analysis. \cite{lo2008agf} have recently developed an
automatic gating approach via robust model-based clustering using
\Rpackage{flowCore}'s data model and infrastructure which is
implemented in the Bioconductor package \Rpackage{flowClust}. Another
package, \Rpackage{plateCore}, providing more specialized support for
experiments conducted on microtitre plates and facilitiating the
handling of spatial metadata, is under development.

@@

\begin{figure}[htbp]
\centering
\includegraphics[width=0.75\textwidth]{flowQ.jpg}
\caption{\label{flowQ}%
  HTML quality assessment report generated by the flowQ package. Rows
  correspond to samples, columns to different quality checks.}
\end{figure}

Static gating for all samples in a high-throughput FCM experiment is
often not an option, since the measured variables tend to vary between
different treatments, over time or between different experiment
batches. Automated or data-driven gating tries to estimate the gating
regions from the underlying data, thus providing a fast objective
solution to the analysis of potentially very large and diverse data
sets \cite{lo2008agf}. One of the automated gating methods implemented
in \Rpackage{flowCore} is based on identifying areas of significant
curvature in a kernel density estimate of the data
\citep{wand2008}. Assuming that the regions of interest are of high
density, the software is able to reliably detect them in a one or
two-dimensional density landscape.

@@

\begin{figure}[htbp]
\centering
\includegraphics[width=0.75\textwidth]{xyplot}
\caption{\label{xyplot}%
  Scatterplot matrix of a single \Rclass{flowSet} with outlines of the
  gating regions identified by an automated gating operation.}
\end{figure}

@@

Through \Rpackage{flowCore}, we have provided the FCM community an open source,
freely available, highly functional, standards compliant development
and analysis platform for high throughput data analysis.  Our
intention is to provide a platform for the collaborative development
of new analysis methods that will facilitate the transition of these
new methods to the larger flow community.  Our experience has been
that the collaborative effort we enable for devising new methodology
has been proven beneficial for a number of different biological and
computational biology challenges.  We hope that our framework will be
the foundation for fruitful shared research by many collaborators from
multiple scientific fields and will help resolve bottlenecks that
currently prevent the further development of and deployment of FC-HCS
to increasingly complex and important scientific and clinical
applications.

>>>>

\subsubsection*{Results}
The \Rpackage{flowCore} package provides the flow cytometry community
with a shared research platform that enables bioinformaticians,
computer scientists, and statisticians to work collaboratively with
biologists and clinicians to develop novel methods for FCM data
analysis.
\subsubsection*{Conclusions}
We hope that our framework will be the foundation for fruitful 
collaborative research involving multiple scientific fields and
that it will help facilitate the
development and deployment of FC-HCS methods to complex 
scientific and clinical applications.

@@

Some of the consequences of this lag in development are difficulties
in maintaining the integrity and documentation of extremely large
datasets, assessing measurement quality, developing validated assays,
controlling the accuracy of gating techniques, automating complex
gating strategies, and aggregating statistical results across large
study sets for further analysis.  Among the most serious consequences
of the current situation is, however, that it is very difficult for
new analysis approaches to find their way into standard practice. We
believe that this barrier to the development and dissemination of new
analysis methods is one of the fundamental restraints on the future
expansion of FC-HCS in both clinical and research applications.

@@

The computational tools we have developed are distributed in the R
software language as the Bioconductor package
\Rpackage{flowCore}. \Rpackage{flowCore} is a freely available, highly
functional and extensible FCM data analysis platform that enables
researchers to efficiently handle FC-HCS data and encourages open
development of tools for their coherent analysis. \Rpackage{flowCore}
implements computationally efficient data structures and a range of
specialized methods addressing all components of a typical FCM
analysis work flow, including compensation, transformation, and
gating. \Rpackage{flowCore} runs on Windows, Mac OS X and Linux/Unix
operating systems. Conceptually, the underlying data structures that
build the core of our software can be implemented in any programming
language. We chose the R language for its advantages in statistical
programming and visualization. In this paper, we describe the
functionality and methods underlying \Rpackage{flowCore}, and provide
examples of their use.

@@

\subsection*{Existing data standards and conventions}
Traditionally, the majority of FCM experiments have been analyzed by
manual data inspection in one or two dimensions, or by very basic
comparisons of summary statistics. Most of the currently available
analysis tools are designed to reflect this workflow.  We believe that
these approaches, in addition to being expensive and labor intensive,
do not fully address the highly complex nature of FCM data; in
particular, they disregard many of the fundamental aspects of the
data, such as its underlying distribution and its high-dimensional
nature. Furthermore, the subjective character of manual analyses are a
major obstacle to reproducibility. In a recent study of flow
cytometric standardization involving 15 institutions, the mean
inter-laboratory coefficient of variation ranged from 17 to 44\%, even
though preparation was standardized and performed using the same
samples and reagents at each site \citep{Maecker2005}. For FC-HCS
data, unassisted manual inspection is extremely time consuming, and
robust statistical methods need to be developed to point investigators
to interesting aspects of the data, or to potential problems. While
the expert knowledge of immunologists and researchers remains crucial
for the understanding of FCM data, we believe that collaboration with
other research fields such as statistics and computer science can
greatly improve the relevance of FCM in today's high-throughput
paradigm.

@@

The notion of classes from an object-oriented programming language
provides one coherent way to describe these richer data structures,
and functions or methods that work on them allow for interaction and
manipulation. Many of the currently available software solutions offer
only limited support for such self-contained structures, or make use
of containers like XML or binary storage formats that are designed
specifically for particular user interfaces and hence not easily
amenable to programmatical access. In addition, the closed-source
nature of these products often makes them impractical to integrate
into analysis pipelines. In this manuscript we describe classes for
FCM data analysis and their implementation in R, however, they could
just as easily be implemented in any other language (e.g., Java,
C++). Software written in those languages could use similar data
structures, thereby simplifying communication and the interchange of
data between analysis tools.

The \Rpackage{flowCore} framework presented here can import and
process raw data FCS files along with their complete set of
file-specific metadata (Figure~\ref{fig1:FrameWork}).  Moreover, it is
a software implementation of the Gating-ML Candidate Recommendation
(CR), an emerging ISAC standard developed in collaboration with the
ISAC Data Standards Task Force, which makes it possibile to integrate
\Rpackage{flowCore} in existing work flows and to communicate with any
other FCM tool that adheres to the proposed standard
\citep{SpidlenInPressCytometryA}. Adherence to standards also plays a
critical role in the ability of new methods based on
\Rpackage{flowCore} to find their way back into the standard practices
for FCM data analysis.

\Rpackage{flowCore} is not a GUI-driven software, and all operations
are done using a command line interface.  It is possible to add a more
elaborate user interface on top of this infrastructure, however the
focus in this paper is on a programmatic approach to enable the
convenient development of novel analysis methods and automation of
complex analysis approaches.  By taking the burden of data management
from the programmer, and by providing well-defined programming
interfaces (APIs), it is possible to readily test new ideas and to
easily extend the framework's functionality.

@@

\Rpackage{flowCore}'s primary task is the representation and basic
manipulation of FCM data. This is accomplished through a data model
very close to that adopted by other Bioconductor packages. All
information from a single FCS file, i.e., the collection of events and
the accompanying metadata, is stored in one single container. We call
the structure that hold this data a \Rclass{flowFrame}. Raw data
values as well as associated metadata of a \Rclass{flowFrame} can be
accessed programmatically. Most commonly, the metadata consist of
descriptors of the stains used in the experiment and the respective
measurement channels, information about compensation performed at the
instrument side and any additional keywords the user deems to be
important to annotate the data. During the creation of a
\Rclass{flowFrame}, a number of quality checks are performed to ensure
data integrity.

\subsubsection*{flowSet}
In high-throughput FCM, many of the analysis tasks need to be
performed consistently across multiple samples, hence we introduce the
concept of a collection of \Rclass{flowFrames} called a
\Rclass{flowSet}. A \Rclass{flowSet} is a container for multiple
\Rclass{flowFrames} along with relevant information associated with
each individual frame such as descriptions of the cell sample, the
treatment to which the sample was subjected, or the location of that
sample in a 96 or 384 well plate.  \Rclass{flowSets} manage the
consistent application of operations on the individual
\Rclass{flowFrames} and shift the burden of keeping score of the
metadata from the user to the infrastructure, thus reducing the risk
of errors (\textit{e.g.}, mixups of sample labels). The
\Rclass{flowSet} structure can be readily extended to incorporate the
potentially complex metadata associated with even larger FC-HCS
experiments such as clinical trials, where hundreds of patients might
provide samples at different time points over the course of the
experiment.

%%Figure1 Diagram of the data structures and the basic operations
\begin{figure}
\centering
\includegraphics[width=0.9\textwidth]{Figure1-flowCoreFrameWork.jpg}
\caption{\label{fig1:FrameWork}{For each experiment, the content of
    the FCS files, phenotypic and metadata are stored in a
    \Rclass{flowSet}. Each \Rclass{flowFrame} in a \Rclass{flowSet}
    corresponds to one FCS file. All basic operations (e.g.,
    compensation, transformation, gating) can be applied to either
    single \Rclass{flowFrames} or a \Rclass{flowSet} simultaneously.}}
\end{figure}

@@

Typically, the basic operations in FCM analyses adhere to the
following common work flow: the data need to be compensated (if that
was not already done on the instrument) and transformed, and
sup-populations of interest need to be selected based on a set of
(predominantly sequential) gates. All software solutions for FCM
analysis offer support for these operations, typically in an
interactive, GUI-driven interface. In \Rpackage{flowCore} we have
taken the approach to abstractly describe these operations and build a
set of tools to perform them on both \Rclass{flowFrames} and
\Rclass{flowSets}. While transformation, and to a certain extent
compensation, are fairly routine operations with only limited
potential for improvement, being able to implement new methodologies
for gating of FCM data, and extending the capabilities of
\Rpackage{flowCore} through object oriented programming are features
that clearly sets our framework apart from other FCM analysis
tools. By factoring out as much of the bookkeeping as possible,
programmers can focus on the actual operations rather then having to
deal with the tedious details of data integration and
access. Third-party methods can act on their own as first-class
citizens in the analysis framework, without breaking the work flow or
the basic infrastructure. This design allows for the straightforward
extension of \Rpackage{flowCore}'s capabilities, and has already
fostered the development of a number of valuable add-ons
\citep{lo2008agf,sarkar2008ufv}.

@@

Data transformation is essential for both data visualization and
modelling \citep{lo2008agf}. The major transformations that are
routinely used in FCM analysis have been implemented in
\Rpackage{flowCore} (\textit{e.g.,} log, bi-exponantioal or arcsinh,
see Table~\ref{table1} for a complete list). Furthermore, the design
of the R language makes it easy to define arbitrary functions to apply
to the data of individual \Rclass{flowFrames} or entire
\Rclass{flowSets}, respectively. Compensation is available for both
\Rclass{flowFrames} and \Rclass{flowSets}, and the software also
offers functionality to compute spillover or compensation matrices
from a set of appropriate single stain controls.

\begin{table}[ht]
\begin{center}
\begin{tabular}{|l|l|}
\hline
\multicolumn{2}{|c|}{Data Transformations} \\
\hline
linear & $ax + b$ \\
quadratic & $ax^2 + bx + c$ \\
natural logarithm & $ln(x)(r/d)$ \\
logarithm & $log(x)(r/d)$ \\
biexponential & $root(a*e^{(b*y)}-c*e^{(-d+y)}+f)$ \\
logicle& $root(Te^{-(m-w)}*(e^{(y-w)}-p^2e^{-(y-w)/p}+p^2-1)$ \\
truncate & $x_{x<=a} = a$ \\
scale & $(x-a)/(b-a)$ \\
split-scale & combination of linear and logarithm \\
arcsinh & $asinh(a + bx)+c$ \\
\hline
\end{tabular}
\caption{\label{table1}Data transformations implemented in
  \Rpackage{flowCore}. Within these formulas, $x$ is the variable
  corresponding to value being transformed, $y$ is the result of the
  transformation, $a$, $b$, $c$, $d$, $f$, $p$, $m$, $T$, and $w$, are
  constants affecting the transformation function, $e$ is the base of
  the natural logarithm, and $root$ is a root-finding algorithm.  }
\end{center}
\end{table}

\subsubsection*{Gating}
In \Rpackage{flowCore}, gating operations are represented by classes
that can be extended in an object-oriented manner
(Table~\ref{table2}). Basic gate types such as rectangular gates,
ellipses and polygon gates are implemented as part of the
framework. In addition, we introduce the notion of data-driven gates,
or filters, for which the necessary parameters are computed based on
the properties of the underlying data, for instance by modelling data
distribution or by density estimation. This approach is fundamentally
different from the traditional application of static gating regions
across samples, as it is able to take into accounts unforeseen changes
in signal intensities, such as drifts in the instrumentation
over time or sample variablity.

@@

Gating, as well as all other operations in \Rclass{flowCore}, can be
applied over each individual frame in a \Rclass{flowSet}, and summary
methods provide information about the outcome of these operations. In
addition, the result of a gating operation can be used to subset the
input \Rclass{flowFrame} or \Rclass{flowSet}, either by filtering out
negative events or by splitting in multiple subpopulations. This
design allows to easily combine all of \Rpackage{flowCore}'s
components into complex work flows.

@@

In addition to the \Rpackage{flowCore} package that offers basic
infrastructure, we have implemented a range of additional Bioconductor
packages that are dedicated to more specific tasks of FCM data
analysis. The \Rpackage{flowViz} package \citep{sarkar2008ufv}
provides sophisticated data visualization tools, that make use off
multivariate trellis plotting \citep{lattice}.  These functions can be
used to quickly generate customized plots for extended cytometry data
sets for both direct data inspection and quality control.  The objects
metadata information can be used to arrange the layout and composition
of the plots.  Furthermore, the design and the API of the
visualization software is very generic, and users can readily extend
its capabilities by providing self-defined plotting functions.  The
\Rpackage{flowQ} package offers more advanced quality assurance (QA)
methodology and a framework to create interactive web-based reports of
QA results. The \Rpackage{flowUtil} package implements data import and
export including flow-cytometry specific standard markup
language. Finally, the \Rpackage{flowStats} package provides elaborate
statistical methods that are relevant in the context of flow
cytometry data analysis. \cite{lo2008agf} have recently developed an
automatic gating approach via robust model-based clustering using
\Rpackage{flowCore}'s data model and infrastructure which is
implemented in the Bioconductor package \Rpackage{flowClust}. Another
package, \Rpackage{plateCore}, providing more specialized support for
experiments conducted on microtitre plates and facilitiating the
handling of spatial metadata, is under development.

@@

\begin{figure}[htbp]
\centering
\includegraphics[width=0.98\textwidth]{flowQ.jpg}
\caption{\label{flowQ}%
  HTML quality assessment report generated by the flowQ package for a
  subset of data from an experiment focussing on Graft-Versus-Host
  Disease \citep{brinkman2007hcf}. Rows correspond to the samples in
  the set, columns to different quality checks.}
\end{figure}

Static gating for all samples in a high-throughput FCM experiment is
often impossible, since the measured variables tend to vary between
different treatments, over time or between different experiment
batches. Automated or data-driven gating has the potential to estimate
the gating regions from the underlying data, thus providing a fast
objective solution to the analysis of potentially very large and
diverse data sets \citep{lo2008agf}. One of the automated gating
methods implemented in \Rpackage{flowCore} is based on identifying
areas of significant curvature in a kernel density estimate of the
data \citep{wand2008}. Assuming that the regions of interest are of
high density, the software is able to reliably detect them in a one or
two-dimensional density landscape.

@@

\begin{figure}[htbp]
\centering
\includegraphics[width=0.75\textwidth]{xyplot}
\caption{\label{xyplot}%
  Scatterplot matrix of a single \Rclass{flowSet} from an experiement
  focussing on immune tolerance following kidney
  transplantation. Outlines of the gating regions identified by an
  automated gating operation are added on top of the density
  representation of the data.}
\end{figure}

@@

Through \Rpackage{flowCore}, we have provided the FCM community an
open source, freely available, highly functional, standards compliant
development and analysis platform for high throughput data analysis.
Our intention is to provide a platform for the collaborative
development of new analysis methods that will facilitate the
transition of these new methods to the larger flow community.  Our
experience has been that the collaborative effort we enable for
devising new methodology has been proven beneficial for a number of
different biological and computational biology challenges.  We hope
that our framework will be the foundation for fruitful shared research
by many collaborators from multiple scientific fields and will help
resolve bottlenecks that currently prevent further development and
deployment of FC-HCS to increasingly complex and important scientific
and clinical applications.
